---
title: "Create occurrence cubes for virtual species"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Apart from simulating occurrence cubes from scratch, **gcube** can also be used to create cubes for virtual species generated using bioclimatic variables. This tutorial demonstrates how to generate a virtual species, apply a geographical limit, sample occurrences, and convert them into an occurrence cube.

# Create virtual species

Virtual species can be created using the **virtualspecies** R package (Leroy et al., [2015](https://doi.org/10.1111/ecog.01388)).
Below is a basic example of how to create a virtual species.
For additional options and a more comprehensive guide, refer to the [package documentation](https://cran.r-project.org/web/packages/virtualspecies/index.html) and the tutorial available [here](https://borisleroy.com/files/virtualspecies-tutorial.html).

```{r setup, warning=FALSE, message=FALSE}
# Load packages
library(gcube)

library(geodata)        # get bioclimatic data
library(virtualspecies) # generate virtual species

library(sf)             # work with spatial objects
library(terra)          # work with raster data
library(ggplot2)        # data visualisation
```

## Step 1: Download WorldClim Data

We first obtain WorldClim bioclimatic data at a 10-minute resolution.

```{r, message=FALSE}
# Download bioclimatic data
worldclim <- worldclim_global(var = "bio", res = 10, path = tempdir())
```

## Step 2: Define the region of interest

We define the extent for the Western Palearctic and crop the climate data.

```{r}
# Western Palearctic
wp <- terra::ext(-15, 65, 30, 75)

# Crop climate data
worldclim <- terra::crop(worldclim, wp)
```

## Step 3: Select relevant bioclimatic variables

We select a subset of bioclimatic variables: bio2, bio5, bio6, bio13, bio14, and bio15.

```{r}
# Specify variables
bio_vars <- c("bio2", "bio5", "bio6", "bio13", "bio14", "bio15")

# Select variables
bio_string <- paste0("bio_", sub("bio", "", bio_vars), collapse = "|")
bio_vars_selected <- names(worldclim)[grepl(bio_string, names(worldclim))]

# Subset climate data
worldclim_vars <- worldclim[[bio_vars_selected]]
```

## Step 4: Generate a virtual species

We create a random virtual species based on the selected bioclimatic variables.

```{r}
#| fig.alt: >
#|   Distribution of random virtual species
set.seed(123)

# Generate random virtual species
virtual_species <- generateRandomSp(worldclim_vars)
```

## Step 5: Apply a geographical limit

We limit the species distribution to the United Kingdom and Ireland.

```{r}
#| fig.alt: >
#|   Limited distribution of random virtual species
# Limit distribution
virtual_distribution <- limitDistribution(
  x = virtual_species,
  geographical.limit = "country",
  area = c("United Kingdom", "Ireland")
)
```

## Step 6: Sample occurrences

We sample 150 occurrences from the restricted distribution with a detection probability of 0.8 using the `virtualspecies::sampleOccurrences()` function.
Detection probability and sampling bias can be introduced similar to `sample_observations()`.

```r
?virtualspecies::sampleOccurrences
```

```{r}
#| fig.alt: >
#|   Samples of random virtual species
# Sample from virtual species distribution
virtual_sample <- sampleOccurrences(
  virtual_distribution,
  n = 150,
  detection.probability = 0.8
)
```

# Grid designation with gcube

Now we can create an occurrence cube for this virtual species
The most important step, is to make the **virtualspecies** output compatible with **gcube** input.
This is done using the `virtualsample_to_sf()` function.

For grid designation, see also the tutorial [3. Specifying the grid designation process](https://b-cubed-eu.github.io/gcube/articles/grid-designation-process.html).

## Step 1: Convert sampled points to sf format

We transform the sampled points from a **virtualspecies** object so that it can be used by **gcube** with the `virtualsample_to_sf()` function.

```r
?virtualsample_to_sf
```

```{r}
# Transform virtual samples to sf object
detections_df_raw <- virtualsample_to_sf(virtual_sample)
detections_df_raw
```

---

*Optional: Extract values from a raster*

> You can also extract raster values when transforming the sample data. For instance, extracting suitability or occurrence probability can be helpful for downstream analysis.

```{r}
# Rename raster layers for clear output
names(virtual_species$suitab.raster) <- "suitability"
names(virtual_species$probability.of.occurrence) <- "occ_prob"

# Transform virtual samples and extract values
virtualsample_to_sf(
  virtual_sample,
  raster_lyr = c(virtual_species$suitab.raster,
                 virtual_species$probability.of.occurrence)
)
```

---

We transform the detected points into an sf object to the LAEA Europe CRS (EPSG:3035).

```{r}
# Transform CRS
detections_df_transformed <- st_transform(detections_df_raw, crs = 3035)
```

## Step 2: Add coordinate uncertainty

See also the tutorial [2. Simulating the detection process](https://b-cubed-eu.github.io/gcube/articles/detection-process.html).

We only keep the detected occurrences using the `filter_observations()` function.

```{r}
detections_df <- filter_observations(detections_df_transformed)
```

Optionally, we can add coordinate uncertainty using the `add_coordinate_uncertainty()` function.
We add 25 m uncertainty for each observation.

```{r}
# Add coordinate uncertainty
observations_df <- add_coordinate_uncertainty(
  observations = detections_df,
  coords_uncertainty_meters = 25
)
```

## Step 3: Specify reference grid

We also need a grid. Each observation will be designated to a grid cell.
You can provide your own grid (e.g. EEA reference grid for Europe).
For this example we create a simple grid around the sampled points.

```{r}
# Create reference grid
buffered_observations <- st_buffer(observations_df, 25)
cube_grid <- st_make_grid(
  buffered_observations,
  n = c(20, 20),
  square = TRUE
) %>%
  st_sf()
```

The grid looks like this.

```{r}
#| fig.alt: >
#|   Grid used for creating data cube.
ggplot() +
  geom_sf(data = observations_df) +
  geom_sf(data = cube_grid, alpha = 0) +
  #coord_sf(datum = 3035) +
  theme_minimal()
```

## Step 4: Perform grid designation

We perform grid designation both in aggregated version and not for visualisation purposes.

```{r}
# Perform grid designation
occurrence_cube_df <- grid_designation(
  observations_df,
  cube_grid,
  seed = 123
)
```

```{r}
# Get sampled points from uncertainty circle
sampled_points <- grid_designation(
  observations_df,
  cube_grid,
  seed = 123,
  aggregate = FALSE
)
```

Lets visualise were the samples were taken.
Note that no distinction is made between zeroes and `NA` values.
Every absence gets a zero value.

```{r}
#| fig.alt: >
#|   Distribution of random samples within uncertainty circle
ggplot() +
  geom_sf(data = occurrence_cube_df,
          alpha = 0) +
  geom_sf_text(data = occurrence_cube_df,
               aes(label = n)) +
  geom_sf(data = sampled_points,
          colour = "blue") +
  #coord_sf(datum = 3035) +
  theme_minimal()
```
