---
title: "Create occurrence cubes for virtual species"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Apart from simulating occurrence cubes from scratch, **gcube** can also be used to create cubes for virtual species generated using bioclimatic variables. This tutorial demonstrates how to generate a virtual species, apply a geographical limit, sample occurrences, and convert them into an occurrence cube.

# Create virtual species

Virtual species can be created using the **virtualspecies** R package (Leroy et al., [2015](https://doi.org/10.1111/ecog.01388)).
Below is a basic example of how to create a virtual species.
For additional options and a more comprehensive guide, refer to the tutorial available [here](https://borisleroy.com/files/virtualspecies-tutorial.html).

```{r setup, warning=FALSE, message=FALSE}
# Load packages
library(gcube)

library(geodata)        # get bioclimatic data
library(virtualspecies) # generate virtual species

library(sf)             # work with spatial objects
library(terra)          # work with raster data
library(ggplot2)        # data visualisation
```

## Step 1: Download WorldClim Data

We first obtain WorldClim bioclimatic data at a 10-minute resolution.

```{r, message=FALSE}
worldclim <- worldclim_global(var = "bio", res = 10, path = tempdir())
```

## Step 2: Define the region of interest

We define the extent for the Western Palearctic and crop the climate data.

```{r}
wp <- terra::ext(-15, 65, 30, 75)
worldclim <- terra::crop(worldclim, wp)
```

## Step 3: Select Relevant Bioclimatic Variables

We select a subset of bioclimatic variables: bio2, bio5, bio6, bio13, bio14, and bio15.

```{r}
bio_vars <- c("bio2", "bio5", "bio6", "bio13", "bio14", "bio15")
bio_string <- paste0("bio_", sub("bio", "", bio_vars), collapse = "|")
bio_vars_selected <- names(worldclim)[grepl(bio_string, names(worldclim))]
worldclim_vars <- worldclim[[bio_vars_selected]]
```

## Step 4: Generate a virtual species

We create a random virtual species based on the selected bioclimatic variables.

```{r}
set.seed(123)
virtual_species <- generateRandomSp(worldclim_vars)
```

## Step 5: Apply a geographical limit

We limit the species distribution to the United Kingdom and Ireland.

```{r}
virtual_distribution <- limitDistribution(
  x = virtual_species,
  geographical.limit = "country",
  area = c("United Kingdom", "Ireland"))
```

## Step 6: Sample occurrences

We sample 150 occurrences from the restricted distribution with a detection probability of 0.8 using the `virtualspecies::sampleOccurrences()` function.
Detection probability and sampling bias can be introduced similar to `sample_observations()`.
See `?virtualspecies::sampleOccurrences` for more options.

```{r}
virtual_sample <- sampleOccurrences(
  virtual_distribution,
  n = 150,
  detection.probability = 0.8)
```

# Grid designation with gcube

See also the tutorial [3. Specifying the grid designation process](https://b-cubed-eu.github.io/gcube/articles/grid-designation-process.html).

## Convert sampled points to sf format

We transform the sample points into an `sf` object:

```{r}
detections_df <- st_as_sf(
  virtual_sample$sample.points,
  coords = c("x", "y"),
  crs = terra::crs(virtual_sample$original.distribution.raster))
```

## Add coordinate uncertainty

Optionally, we can add coordinate uncertainty using the `add_coordinate_uncertainty()` function.
See also the tutorial [2. Simulating the detection process](https://b-cubed-eu.github.io/gcube/articles/detection-process.html).
We add 25 m uncertainty for each observation.

```{r}
observations_df <- add_coordinate_uncertainty(
  observations = detections_df,
  coords_uncertainty_meters = 25)
```

## Specify reference grid

We also need a grid. Each observation will be designated to a grid cell.
You can provide your own grid (e.g. EEA reference grid for Europe).
For this example we create a simple grid around the sampled points.

```{r}
buffered_observations <- st_buffer(observations_df, 25)
cube_grid <- st_make_grid(
  buffered_observations,
  n = c(20, 20),
  square = TRUE) %>%
  st_sf()
```

The grid looks like this.

```{r}
#| fig.alt: >
#|   Grid used for creating data cube.
ggplot() +
  geom_sf(data = observations_df) +
  geom_sf(data = cube_grid, alpha = 0) +
  theme_minimal()
```

## Perform grid designation

```{r}
occurrence_cube_df <- grid_designation(
  observations_df,
  cube_grid,
  seed = 123)
```

```{r}
sampled_points <- grid_designation(
  observations_df,
  cube_grid,
  seed = 123,
  aggregate = FALSE)
```

Lets visualise were the samples were taken for time point 1.
Note that no distinction is made between zeroes and `NA` values!
Every absence gets a zero value.

```{r}
#| fig.alt: >
#|   Distribution of random samples within uncertainty circle
ggplot() +
  geom_sf(data = occurrence_cube_df,
          alpha = 0) +
  geom_sf_text(data = occurrence_cube_df,
               aes(label = n)) +
  geom_sf(data = sampled_points,
          colour = "blue")
  theme_minimal()
```
