---
title: "Grid designation for custom datasets"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial explains how spatial grid designation is performed in ecological datasets using the **gcube** package. We begin with the theoretical foundations of generating random points within uncertainty buffers and follow with a practical example of applying a uniform grid to a real dataset. The method supports both projected and geographic coordinate reference systems, making it applicable beyond simulated data.

## Method

To generate a point within a circle of radius $r$ around a central coordinate $(x_0, y_0)$, we use polar coordinates and convert to Cartesian coordinates.

1. Sample a **random angle** from the uniform distribution $\mathcal{U}$:
   
$$
\theta \sim \mathcal{U}(0, 2\pi)
$$

2. Sample a **random radius** scaled by square root to ensure uniform point density:

$$
r' = \sqrt{\mathcal{U}(0, 1)} \cdot r
$$

3. Convert to Cartesian coordinates:

$$
x = x_0 + r' \cos(\theta)
$$

$$
y = y_0 + r' \sin(\theta)
$$

## Coordinate Reference System Considerations

The method of applying the above formula depends on the unit system of your CRS.

### Projected CRS (e.g., UTM)

Projected systems use **meters** or other linear units, so the polar formula can be applied directly.

| Quantity | Formula |
|----------|---------|
| $x$ offset | $x = x_0 + r \cos(\theta)$ |
| $y$ offset | $y = y_0 + r \sin(\theta)$ |

This is the most accurate and straightforward method and should be used whenever your data are in a projected CRS.

### Geographic CRS (e.g., WGS84)

Geographic systems use **degrees**, not meters. Therefore, the uncertainty radius in meters must be converted to degrees using approximations based on latitude:

- 1 degree of latitude ≈ 111,320 meters
- 1 degree of longitude ≈ $111{,}320 \cdot \cos(\text{latitude})$ meters

Use these formulas to convert meters to degrees:

| Quantity | Formula |
|----------|---------|
| Latitude offset | $\Delta \text{lat} = \dfrac{r \sin(\theta)}{111320}$ |
| Longitude offset | $\Delta \text{lon} = \dfrac{r \cos(\theta)}{111320 \cdot \cos(\text{lat})}$ |

These are **approximations** and may introduce distortion, especially at high latitudes.

> ⚠️ **Tip**: Always prefer using a projected CRS (like UTM) for grid designation when available. This minimizes distortion and simplifies calculations. Use geographic CRS (e.g., WGS84) only if your dataset explicitly requires it or no projected CRS is available.

## Getting started with gcube

In this example, we show how **gcube** can be used to perform grid designation on a custom ecological dataset.

```{r, message=FALSE, warning=FALSE}
# Load packages
library(gcube)


library(frictionless) # Load example dataset
library(ggplot2)      # Data visualisation
library(dplyr)        # Data wrangling
library(sf)           # Working with spatial data
```

### Get example data

As reference grid, we load the the MGRS grid at 10 km scale from the **b3data** data package.

```{r}
# Read data from repository
mgrs10_belgium <- st_read(
  "https://zenodo.org/records/15211029/files/mgrs10_refgrid_belgium.gpkg",
  quiet = TRUE
)
head(mgrs10_belgium)
```

Consider the following synthetic observation dataset with three species observed across three different years. Each row represents a record with spatial coordinates and an associated uncertainty in meters.
Both the reference grid and observation data use the same projected CRS, which ensures that distances and buffer sizes are accurately computed.

```{r}
# Simulate 9 observation records: 3 species x 3 years
example_data <- data.frame(
  species = rep(c("species1", "species2", "species3"), each = 3),
  time_point = rep(2020:2022, times = 3),
  lat = c(
    5683521, 5613261, 5533663,
    5672768, 5559184, 5670045,
    5628962, 5651001, 5587280
  ),
  long = c(
    512309.4, 560722.6, 700880.4,
    649960.8, 688510.1, 641283.4,
    502044.7, 568593.9, 566011.5
  ),
  coordinateUncertaintyInMeters = c(
    1000, 2500, 5000,
    1500, 10000, 8000,
    4000, 2500, 3500
  )
) %>%
  st_as_sf(coords = c("long", "lat"), crs = st_crs(mgrs10_belgium))

example_data
```

We visualise the data with uncertainty circles.

```{r}
#| fig.alt: >
#|   Visualisation of example dataset
# Create sf object with uncertainty circles
buffered_observations <- st_buffer(
  example_data,
  example_data$coordinateUncertaintyInMeters
)

# Visualise
ggplot() +
  geom_sf(data = mgrs10_belgium) +
  geom_sf(data = buffered_observations,
          fill = alpha("firebrick", 0.3)) +
  geom_sf(data = example_data, aes(colour = species))
```

### Grid designation

To perform grid designation, we iterate over each species, filter the corresponding observations, and apply `grid_designation()` to generate a species-specific data cube.  
For improved performance, especially with larger datasets, consider using vectorized approaches such as `lapply()` or `purrr::map()` instead of explicit loops.

```{r}
# Get species
taxa <- sort(unique(example_data$species))

# Create empty list
occurrence_cube_list <- vector(mode = "list", length = length(taxa))

# Loop over species
for (i in seq_along(taxa)) {
  # Get species
  taxon <- taxa[i]

  # Filter data
  taxon_data <- example_data %>%
    filter(taxon == species)

  # Perform grid designation
  taxon_cube <- grid_designation(
    observations = example_data,
    grid = mgrs10_belgium,
    seed = 123
  )

  # Add species column
  taxon_cube$species <- taxon

  # Add species cube to list
  occurrence_cube_list[[i]] <- taxon_cube
}

# Combine species cubes
occurrence_cube_full <- bind_rows(occurrence_cube_list)
```

This will however result in many duplicate rows for grid cells where no species were seen.
They can be removed by filtering or unique rows can be retained using the `dplyr::distinct()` function.

```{r}
start_time <- Sys.time()

occurrence_cube_df1 <- distinct(occurrence_cube_full)

end_time <- Sys.time()
end_time - start_time
```

Geometry columns are often large and complex, making operations like `distinct()` significantly slower.
Dropping geometry before deduplication and rejoining spatial info afterwards drastically improves performance on large datasets.
A faster option is to drop the geometry and join it again with the grid after retaining unique columns.

```{r}
start_time <- Sys.time()

occurrence_cube_df2 <- occurrence_cube_full %>%
  st_drop_geometry() %>%
  distinct() %>%
  left_join(mgrs10_belgium, by = join_by(mgrscode)) %>%
  st_sf(crs = st_crs(mgrs10_belgium))

end_time <- Sys.time()
end_time - start_time
```

```{r}
identical(occurrence_cube_df1, occurrence_cube_df2)
```

We visualise the final occurrence cube.
Since we have only three species occurrences per year spread over the whole grid, we see three occupied grid cells per year.

```{r}
#| fig.alt: >
#|   Visualisation of example data cube
ggplot(occurrence_cube_df1) +
  geom_sf(aes(fill = factor(n))) +
  facet_wrap(~time_point, nrow = 2) +
  labs(fill = "number of\nobservations")
```
