---
title: "4. Creating cubes for multiple species"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The workflow for simulating a biodiversity data cubes is provided in tutorials 1-3.
However, this workflow only describes a data cube for a single species.
As such, it is not really a cube since it is not three-dimensional.
Furthermore, many biodiversity indicators are calculated over multiple species.

This tutorial describes the functions provided to simulate biodiversity data cubes for multiple species directly.

```{r setup, warning=FALSE, message=FALSE}
# Load packages
library(gcube)

library(sf)        # work with spatial objects
library(dplyr)     # data wrangling
library(ggplot2)   # data visualisation
```

# Spatial extend

As input, we create a polygon in which we simulate occurrences.

```{r}
polygon <- st_polygon(list(cbind(c(500, 1000, 1000, 600, 200, 100, 500),
                                 c(200, 100, 700, 1000, 900, 500, 200))))
```

The polygon looks like this.

```{r}
#| fig.alt: >
#|   Spatial extend in which we will simulate species occurrences.
ggplot() +
  geom_sf(data = polygon) +
  theme_minimal()
```

Also consider a road across our polygon.

```{r}
# Define the road width
road_width <- 50

# Create road points
road_points <- rbind(c(100, 500), c(1000, 500))

# Create road-like polygon within the given polygon
road_polygon <- st_linestring(road_points) %>%
  st_buffer(road_width) %>%
  st_intersection(polygon) %>%
  st_polygon() %>%
  st_sfc() %>%
  st_as_sf() %>%
  rename(geometry = x)
```

The result looks like this.

```{r}
#| fig.alt: >
#|   Spatial extend with road in which we will simulate species occurrences.
ggplot() +
  geom_sf(data = polygon, fill = "lightgreen") +
  geom_sf(data = road_polygon) +
  theme_minimal()
```

We also need a grid.
Each observation will be designated to a grid cell.

```{r}
cube_grid <- st_make_grid(
  st_buffer(polygon, 50),
  n = c(20, 20),
  square = TRUE) %>%
  st_sf()
```

The grid looks like this.

```{r}
#| fig.alt: >
#|   Grid used for creating data cube.
ggplot() +
  geom_sf(data = polygon) +
  geom_sf(data = cube_grid, alpha = 0) +
  theme_minimal()
```

# Input dataframe

To generate cubes for multiple species efficiently, we create a dataframe where each row represents a different species and where we specify all arguments to be used by the main cube simulation functions, viz `simulate_occurrences()`, `sample_observations()`, `filter_observations()`, `add_coordinate_uncertainty()`, and `grid_designation()`, in separate columns.
The values within these columns can change between species.

```{r}
# Create dataframe with simulation function arguments
multi_species_args <- tibble(
  species_range = rep(list(polygon), 6),
  initial_average_occurrences = rep(c(50, 100, 500), 2),
  n_time_points = rep(6, 6),
  temporal_function = c(simulate_random_walk, simulate_random_walk, rep(NA, 4)),
  sd_step = c(1, 1, rep(NA, 4)),
  spatial_autocorr = c(rep("random", 3), rep("clustered", 3)),
  detection_probability = rep(c(0.8, 0.9, 1), 2),
  sampling_bias = c(rep("polygon", 3), rep("no_bias", 3)),
  bias_area = c(rep(list(road_polygon), 3), rep(NA, 3)),
  bias_strength = c(c(0.1, 0.2, 0.3), rep(1, 3)),
  coords_uncertainty_meters = rep(c(25, 30, 50), 2),
  grid = rep(list(cube_grid), 6),
  seed = 123
)

# How does this dataframe look like?
glimpse(multi_species_args)
```

Before we move on, we add a taxonomic hierarchy to this dataframe.
This might be used to calculate phylogenetic indicators upstream or just to have a more complete and comprehensible dataset for cube generation.
We generate a taxonomic hierarchy using the `generate_taxonomy()` function.

```{r, eval=FALSE}
?generate_taxonomy
```

We need to specify at least the number of species, genera and families.
Higher taxonomic levels are default on 1.

```{r}
# Define number of species
n_specs <- nrow(multi_species_args)

# Generate taxonomic hierarchy
taxonomy <- generate_taxonomy(
  num_species = n_specs,
  num_genera = 4,
  num_families = 2,
  seed = 123)

# How does the taxonomic hierarchy look like?
taxonomy
```

We can add this taxonomic hierarchy to our dataframe with function arguments to get a complete dataset.

```{r}
# Combine dataframes
multi_species_dataset1 <- cbind(taxonomy, multi_species_args) %>%
  as_tibble() # To compare with dataset2, see next chunk

# How does the dataset look like?
glimpse(multi_species_dataset1)
```

Or we can give the dataframe directly as input to `generate_taxonomy()`, which will create a species with taxonomic hierarchy for each row.

```{r}
# Use dataframe as input to generate taxonomic hierarchy
multi_species_dataset2 <- generate_taxonomy(
  num_species = multi_species_args,
  num_genera = 4,
  num_families = 2,
  seed = 123)

# Do we get the same result?
identical(multi_species_dataset1, multi_species_dataset2)
```

# Mapping the simulation process over each row/species

Each cube simulation function has a mapping companion.
These mapping functions apply the single-species operations for each row using the `purrr::pmap()` strategy.

```{r, echo=FALSE}
data.frame(
    single_species = c("simulate_occurrences()",
                       "sample_observations()",
                       "filter_observations()",
                       "add_coordinate_uncertainty()",
                       "grid_designation()"),
    multi_species  = c("map_simulate_occurrences()",
                       "map_sample_observations()",
                       "map_filter_observations()",
                       "map_add_coordinate_uncertainty()",
                       "map_grid_designation()")
  ) %>% knitr::kable()
```

The output of the mapping functions is stored as a list-column by default to ensure efficient transition between simulation functions.

```{r}
# Creates occurrences list-column with simulated occurrences
map_sim_occ_df <- map_simulate_occurrences(df = multi_species_dataset2)

# Creates observations_total list-column with sampled observations
map_samp_obs_df <- map_sample_observations(df = map_sim_occ_df)

# Creates observations list-column with detected observations
map_filter_obs_df <- map_filter_observations(df = map_samp_obs_df)

# Add coordinateUncertaintyInMeters column to observations list-column
map_obs_uncertainty_df <- map_add_coordinate_uncertainty(df = map_filter_obs_df)

# Creates occurrence_cube_df list-column with final data cubes
map_occ_cube_df1 <- map_grid_designation(df = map_obs_uncertainty_df)

# How does the final dataframe look like?
glimpse(map_occ_cube_df1)
```

We can also pipe over the functions.

```{r}
map_occ_cube_df2 <- multi_species_dataset2 %>%
  map_simulate_occurrences() %>%
  map_sample_observations() %>%
  map_filter_observations() %>%
  map_add_coordinate_uncertainty() %>%
  map_grid_designation()

# Is the output the same as before?
identical(map_occ_cube_df1, map_occ_cube_df2)
```

Each mapping function also has an argument `nested`.
If `nested = TRUE` (default), they retain the list-column containing calculated sf objects calculated by simulate_occurrences().
If `nested = TRUE`, this list-column is expanded into rows and columns.

If we do this for example for `map_grid_designation()`, we can recognise the columns produced by `grid_designation()`.
Nested list-columns are useful for calculation in the simulation workflow as shown above.
Calculation and visualisation of the output is more easily performed on unnested list-columns.

```{r}
# Unnest list-column
map_occ_cube_df_unnested <- map_grid_designation(
  df = map_obs_uncertainty_df,
  nested = FALSE)

# How does the dataset look like?
glimpse(map_occ_cube_df_unnested)
```

Finally, the mapping functions allow column names that deviate from the argument names of their single-species companions.
A names list should then be provided to the mapping functions to know which columns match which arguments.
The mapping functions will retain the original column names in the output.

We rename for example three column names such that they do not longer match the argument names of the single-species functions:

- species_range $\rightarrow$ polygon (argument of `simulate_occurrences()`)
- detection_probability $\rightarrow$ det_prob (argument of `sample_observations()`)
- grid $\rightarrow$ raster (argument of `grid_designation()`)

```{r}
# Create dataframe with custom column names
multi_species_dataset2_renamed <- multi_species_dataset2 %>%
  rename("polygon" = "species_range",
         "det_prob" = "detection_probability",
         "raster" = "grid")
```

We need to create a named list to match the new column names with the argument names.
The names are the argument names of the single-species functions and the values are the deviating column names.

```{r}
# Create named list for argument conversion
arg_conv_list <- list(
    species_range = "polygon",
    detection_probability = "det_prob",
    grid = "raster"
  )

arg_conv_list
```

We perform the same workflow as before and use the `arg_list` argument convert the names.
We can see that the workflow works as expected and the column names are the original (changed) names of the input dataframe.

```{r}
map_occ_cube_df3 <- multi_species_dataset2_renamed %>%
  map_simulate_occurrences(arg_list = arg_conv_list) %>%
  map_sample_observations(arg_list = arg_conv_list) %>%
  map_filter_observations() %>% # Argument names match
  map_add_coordinate_uncertainty() %>% # Argument names match
  map_grid_designation(arg_list = arg_conv_list)

# How does the dataset look like?
glimpse(map_occ_cube_df3)
```

# Visualise examples

Let's visualise the output for two of the six species.

Species 1 has a low initial number of occurrences with a random walk over time and is simulated using a random spatial pattern.
It has a detection probability of 0.8 and will be undersampled on the road.

```{r}
# Simulate detection process and unnest
detections_df_raw <- map_sample_observations(df = map_sim_occ_df,
                                             nested = FALSE)
# Select species 1 and create sf object
detections_df_raw1 <- detections_df_raw %>%
  filter(species == "species1") %>%
  st_sf()
```

We confirm this by visualising the spatial distribution of the occurrences with their sampling status.

```{r}
#| fig.alt: >
#|   Spatial distribution of occurrences with indication of sampling status for
#|   each time point.
ggplot() +
  geom_sf(data = polygon, fill = "lightgreen") +
  geom_sf(data = road_polygon) +
  geom_sf(data = detections_df_raw1,
          aes(colour = sampling_status)) +
  scale_colour_manual(values = c("blue", "red")) +
  facet_wrap(~time_point, nrow = 2) +
  labs(title = "Distribution of occurrences for each time point") +
  theme_minimal()
```

We do the same for species 6.
Species 6 has a high initial number of occurrences with no trend over time and is simulated using a clustered spatial pattern.
It has a detection probability of 1 and there is no sampling bias.

```{r}
#| fig.alt: >
#|   Spatial distribution of occurrences with indication of sampling status for
#|   each time point.
# Select species 6 and create sf object
detections_df_raw6 <- detections_df_raw %>%
  filter(species == "species6") %>%
  st_sf()

ggplot() +
  geom_sf(data = polygon, fill = "lightgreen") +
  geom_sf(data = road_polygon) +
  geom_sf(data = detections_df_raw6,
          aes(colour = sampling_status)) +
  scale_colour_manual(values = c("blue", "red")) +
  facet_wrap(~time_point, nrow = 2) +
  labs(title = "Distribution of occurrences for each time point") +
  theme_minimal()
```
